---
title: "Reproducible Research - Course Project 1"
author: "Thomas Hill III"
date: "February 26, 2019"
output: html_document
---
```{r setoptions, echo = TRUE}
```
##Loading and preprocessing the data  
Show any code that is needed to  
1.) Load the data (i.e. read.csv())  
Note that setting the working directory may or may not be necessary depending on your directory setup.
```{r}
#setwd("./Reproducible Research")
```
The code below includes a way to measure the time it took to read in the data. You don't have to include the a,b, or b-a code in your file. It makes no difference to anything that comes after. It's just interesting to see.
```{r}
a <- Sys.time(); act <- read.csv("activity.csv", as.is = TRUE); b <- Sys.time(); b-a
```
2.) Process/transform the data (if necessary) into a format suitable for your analysis  
[I don't do anything here. All of the data processing is done within the actual answers]

##What is the mean total number of steps taken per day?  
For this part of the assignment, you can ignore the missing values in the dataset.  
1.) Calculate the total number of steps taken per day  
```{r results = "hide", message=FALSE, warning = FALSE}
library(dplyr)
library(lubridate)
```
```{r}
act1 <- 
        act %>%
        mutate(month = month(ymd(date))
               , day = day(ymd(date))) %>%
        group_by(month, day) %>%
        summarise(sumSteps = sum(steps))
act1
```
2.) Make a histogram of the total number of steps taken each day  
```{r}
par(cex = .7)
hist(act1$sumSteps
     , col = "green"
     , main = "Histogram of Total Number of Steps Taken Each Day (act1)"
     , xlab = "Number of Steps"
     , ylab = "Number of Days With That Many Steps"
     , xaxt = "n"
     , breaks = 61
)
axis(side = 1
     , at = seq(from = 0, to = 25000, by = 1000)
     , las = 2)
```

3.) Calculate and report the mean and median of the total number of steps taken per day  
```{r}       
act2 <- act %>%
        mutate(
                year = year(ymd(date))
                ,
                month = month(ymd(date))
                ,
                day = day(ymd(date))
                ,
                monthDay = as.Date(paste(year(ymd(date)),"-", month,"-", day, sep = ""))
        ) %>%
        group_by(monthDay) %>%
        summarise(
                meanSteps = mean(steps, na.rm = TRUE),
                medianSteps = median(steps, na.rm = TRUE)
        )
```
This is the mean data
```{r}
a <- !is.na(act2$meanSteps)
act2$meanSteps[a]
```

3.) continued - For an over-time graph this is best displayed in a lineplot  
```{r}

plot(x = act2$monthDay[a]
     ,y = act2$meanSteps[a]
     ,type = "l"
     , col = "darkred"
     , ylab = "Mean Number of Steps in the Day"
     , xlab = "Date (in 2012)"
     , main = "Mean Number of Steps Taken Each Day" 
     , xaxt = "n"
     , xlim = c(as.Date("2012-10-01"), as.Date("2012-11-30"))
)
axis(side = 1, at = act2$monthDay[seq(from = 1, to = 61, by = 5)], format(act2$monthDay[seq(from = 1, to = 61, by = 5)], "%b %d"), las = 2, cex.axis = .7)
```
  
This is the median data
```{r}
act2$medianSteps[a]
```
```{r}
plot(x = act2$monthDay[a]
     , y = act2$medianSteps[a]
     , type = "l"
     , col = "darkslateblue"
     , ylab = "Median Number of Steps in the Day"
     , xlab = "Date (in 2012)"
     , main = "Median Number of Steps Taken Each Day" 
     , xaxt = "n"
     , xlim = c(as.Date("2012-10-01"), as.Date("2012-11-30"))
     , ylim = c(-.1, 35)
)
axis(side = 1, at = act2$monthDay[seq(from = 1, to = 61, by = 5)], format(act2$monthDay[seq(from = 1, to = 61, by = 5)], "%b %d"), las = 2, cex.axis = 0.6)
```

##What is the average daily activity pattern?  
1.) Make a time series plot (i.e. type = "l")  of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).  
```{r}
act3 <- act %>%
        group_by(interval) %>%
        summarise(
                meanStepsTI = mean(steps, na.rm = TRUE)
        )
act3
```
```{r}
plot(x = act3$interval[a]
     ,y = act3$meanStepsTI[a]
     ,type = "l"
     , col = "black"
     , ylab = "Mean Number of Steps in Each Time Interval"
     , xlab = "Time Interval"
     , main = "Average Number of Steps Taken in Each Time Interval"
     , xaxt = "n"
     , xlim = c(min(act3$interval[a]), max(act3$interval[a]))
)
axis(side = 1, at = act3$interval[rep(c(T,F,F,F,F,F,F,F,F,F,F,F), 24)], las = 2, cex.axis = .7)
```
       
2.) Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
```{r results = "hide"}
maxMeanStepsTI <- max(y = act3$meanStepsTI[a])
maxMeanStepsTI #206.1698
maxStepsIntervalIndex <- which(act3$meanStepsTI == maxMeanStepsTI)
maxStepsIntervalIndex
```
```{r}
act3$interval[maxStepsIntervalIndex]
```
####So the 835-840 interval has the greatest average number of steps across all the days.  
  
##Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.  
 
1.) Calculate and report the total number of missing values in the dataset (i.e. # total number of rows with NAs).  
```{r}
sum(is.na(act$steps))
sum(is.na(act$date))
sum(is.na(act$interval))
```
There are 2304 missing values, and they are all in "steps". Note that 2304 = 8 days * 288 intervals/day.  

2.) Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.  
   
Personal note: I could use a simple mean here.  
*a.) Would a linear regression over each day make any sort of sense? It might, but coding it is going to be a mite complicated.  
*b.) perhaps some sort of simple linear interpolation? But not all the missing values are flanked by present values.  
*c.) Some days are missing data completely, so we can't use a daily average.  
*d.) Perhaps the simplest (and most comprehensive) thing to do then is to use the multi-day average for the specific interval.  
  
3.) Create a new dataset that is equal to the original dataset but with the missing data filled in.  
```{r results = "hide"}
act5 <- act
act4 <- act %>% filter(is.na(steps))
dim(act4) #2304x3


for (interv in act4$interval) {
        indexList <- grep(interv, act4$interval)
        act4$steps[indexList] <- act3$meanStepsTI[indexList[1]]
}
```
We can check our work.  
```{r}
head(act4)
```
Great. Looks like we've got the new data. Now we just have to put it into the main dataframe.  
```{r}
for (i in 1:length(act$steps)) {
        if (is.na(act5$steps[i])) {
                act5$steps[i] <- act4$steps[(i %% 288)+1]
        }
}
sum(is.na(act5$steps))
```
Looks good.  
This means that there are no missing values in the steps column any longer.  
  
We can check our work -  
```{r}
head(act5); 
```
```{r}
head(act)
```
Note that these show the same identifying columns and the NAs have been replaced.
That's good. That's what we wanted.  
  
4.a.)Make a histogram of the total number of steps taken each day  
```{r}
library(dplyr)
library(lubridate)
act6 <- act5 %>%
        mutate(
                month = month(ymd(as.character(date)))
                ,
                day = day(ymd(date))
                ,
                year = year(ymd(date))
                ,
                yearMonthDay = as.Date(paste(year, "-", month, "-", day, sep = ""))
        ) %>%
        group_by(yearMonthDay) %>%
        summarise(
                sumStepsTI = sum(steps),
                meanStepsTI = mean(steps),
                medianStepsTI = median(steps)
        )
act6
```
```{r}
par(mar = c(4,2,3,1))
hist(act6$sumStepsTI
     , col = "green"
     , main = "Histogram of Number of Daily Steps (act6)"
     , xlab = "Number of Steps"
     , ylab = "Number of Days With That Many Steps"
     , xaxt = "n"
     , breaks = 61
     , ylim = c(0,7)
)
mtext("(Y axis rescaled to match the first histogram)")
axis(side = 1
     , at = seq(from = 0, to = 25000, by = 1000)
     , las = 2
)
```
  
4.b.)Calculate and report the mean and median total number of steps taken per day.  
This is the mean data
```{r}
act6$meanStepsTI
```
  
This is a graph of the mean data.
```{r}
plot(
        x = act6$yearMonthDay
        ,y = act6$meanStepsTI
        ,col = "darkred"
        ,type = "l"
        ,pch = floor(runif(1, 1, 26))
        ,ylab = "Mean Number of Steps"
        ,main = "Mean Number of Steps in the Day"
        ,xlab = "Date"
        ,xaxt = "n"
        ,xlim = c(as.Date("2012-10-01"), as.Date("2012-11-30"))
)
labels <- act6$yearMonthDay[seq(from = 1, to = 61, by = 5)]
axis(side = 1, at = labels, format(labels, "%b %d"), las = 2, cex.axis = .7)
```
  
This is a graph of the median data.
```{r}
plot(
        x = act6$yearMonthDay
        ,y = act6$medianStepsTI
        ,col = "darkslateblue"
        ,ylab = "Median Number of Steps"
        ,main = "Median Number of Steps in the Day"
        ,xlab = "Date"
        ,xaxt = "n"
        ,xlim = c(as.Date("2012-10-01"), as.Date("2012-11-30"))
        ,ylim = c(-.1, 35)
)
axis(side = 1, at = labels, format(labels, "%b %d"), las = 2, cex.axis = 0.6)
```


  
Do these values differ from the estimates from the first part of the assignment?   
 *-Histogram is mostly the same except for a jump in the 10500-11000.   
 *-Median graph just shows some non-zero values.  
 -Mean graph shows very little change, but there is some.  

What is the impact of imputing missing data on the estimates of the total daily number of steps?  
 -It doesn't change much. We see a single large jump for the 10500-11000 interval and a few other minor increments in other bars.  

##Are there differences in activity patterns between weekdays and weekends?  
1.) Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.  
```{r}
library(lubridate)
is.weekend <- function(input) {
        input <- as.Date(input)
        if (wday(input) %in% c(1, 7)) {
                return("weekend")
        } else {
                return("weekday")
        }
}
table(sapply(act$date, is.weekend))
```
See that there are 12960 of weekday and 4608 of weekend.  
Note that 4608 = 8 weekends * 2 days/weekend * 288 intervals/day.  
```{r}
act7 <- act5
act7$WKND <- factor(sapply(act$date, is.weekend))
```
  
Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).  
```{r}
act8 <- act7 %>%
        group_by(interval, WKND) %>%
        summarise(
                meanStepsTI = mean(steps, na.rm = TRUE)
        )
act8
```
```{r}
library(lattice)
xyplot(act8$meanStepsTI ~ act8$interval |
               WKND
       ,data = act8
       ,xlab = "Interval Start"
       ,ylab = "Mean Steps"
       ,type = "l"
       ,main = "Mean Number of Steps per Interval"
       ,layout = c(1,2))
```